---
title: "Estimands: A Working Primer"
subtitle: "Seeing them, writing them, and critiquing them in real research"
engine: knitr
---

```{=html}
<!-- ============ Quiz engine: client-side, no server, survives embed-resources ============ -->
<script>
function estCheck(id){
  const q = document.getElementById(id);
  const chosen = q.querySelector('input[type=radio]:checked');
  const fb = q.querySelector('.est-feedback');
  if(!chosen){ fb.innerHTML = '<em>Pick an answer first.</em>'; fb.className = 'est-feedback show neutral'; return; }
  const ok = chosen.getAttribute('data-correct') === 'true';
  const why = chosen.getAttribute('data-why') || '';
  fb.innerHTML = (ok ? '✅ <strong>Correct.</strong> ' : '✗ <strong>Not quite.</strong> ') + why;
  fb.className = 'est-feedback show ' + (ok ? 'ok' : 'no');
  // mark options
  q.querySelectorAll('label.est-opt').forEach(l => l.classList.remove('picked-ok','picked-no'));
  chosen.closest('label').classList.add(ok ? 'picked-ok' : 'picked-no');
}
</script>
```

::: {.callout-note appearance="simple"}
This is a self-contained learning chapter. The multiple-choice checkpoints run
entirely in your browser, and every R code block shows its computed output
inline, so the whole document works offline with no dependencies.
:::

# Orientation {#sec-orientation}

This chapter is about one deceptively small word. An **estimand** is *the thing
you are trying to learn* — the quantity that would answer your question if you
had infinite data and no estimation problems. It is not a method, not a
coefficient, and not a number you report. It is the target: the precise,
answerable form of your **research question** — the hinge between a sentence in
your introduction (*"does X affect Y, and for whom?"*) and a quantity a method
can actually chase.

The trouble is that in clean textbook cases the target is obvious, and in real
political-science cases it is often not. The same design can carry several
plausible estimands; methods quietly substitute one target for another; and a
paper can state one estimand in its design section and defend a different one in
its conclusion. The cost of fuzziness here is not pedantic. It is the difference
between *"this program works"* and *"this program worked for the people who were
already getting it,"* which are different claims with different policy
implications.

## Who this is for

You are comfortable with research design, causal inference, and estimation. You
can already define an ATE, you know an estimator from an estimate, you have seen
LATE and the potential-outcomes notation. **This chapter does not re-teach any
of that as new material.** It treats the definitions as a *fast recap* and
spends its real estate on the part that stays hard:

- using the taxonomy as working vocabulary, not flashcards;
- reasoning *from a messy design to its actual estimand*;
- spotting where intuition fails and the target quietly shifts;
- writing a clean estimand sentence in a paper;
- critiquing other people's estimands out loud, in workshop and referee language.

## How to read it

Read @sec-recap and @sec-what fast — they set notation and the Lundberg–Johnson–Stewart
(LJS) anatomy you will reuse everywhere. @sec-fieldguide and @sec-designmap are the
**reference** sections — the taxonomy of estimands and the map from designs to the
estimands they target; skim them now, return later. The heart of the chapter is
@sec-breaks (where intuition breaks) through @sec-critique (critiquing others), with
@sec-sampling adding the population question most survey work leaves implicit.
@sec-cad applies everything to a real coastal-flooding study — skip it if you are
an external reader.

::: {.callout-important appearance="simple"}
**The point is not more notation. It is clearer targets.** If you finish this
chapter able to recite definitions but unable to say, in a seminar, *"I think
the design identifies X but the paper is written as if it identifies Y,"* the
chapter has failed.
:::

# Fast recap: estimand, estimator, estimate {#sec-recap}

Three words that sound similar and are constantly conflated. Keep them in three
different drawers.

| Term | What it is | Mean example | Lives where |
|---|---|---|---|
| **Estimand** | The quantity you want to know | The population mean income, $\mu = E[Y]$ | In the world (a fact) |
| **Estimator** | The *rule/procedure* for guessing it | "Take the sample average," $\bar Y = \tfrac1n\sum Y_i$ | In your method |
| **Estimate** | The *realised number* the rule produces | $\bar y = 41{,}300$ kr. | In your output table |

: The three drawers {.striped}

The estimand exists whether or not you ever collect data. The estimator is a
function of data you have not yet seen. The estimate is what you get once you
plug the data in — a single draw from the estimator's sampling distribution. A
good estimator is one whose estimates cluster around the estimand; "unbiased,"
"consistent," and "efficient" are all statements about *estimator behaviour
relative to a fixed estimand*. None of them mean anything until the estimand is
pinned down.

```{r}
#| label: demo-mean
#| message: false
set.seed(2026)

# The ESTIMAND: a population mean we are pretending to chase.
mu <- 41300                      # true mean income (kr.), unknown in real life

# Draw one sample from that population.
pop_sd <- 9000
y <- rnorm(40, mean = mu, sd = pop_sd)

# The ESTIMATOR: the sample-mean rule.  The ESTIMATE: the number it returns.
estimate <- mean(y)

cat("Estimand (truth):", mu, "\n")
cat("Estimate (this sample):", round(estimate), "\n")
cat("Estimation error:", round(estimate - mu), "\n")
```

The *estimand* `mu` never moves; the *estimate* jumps around it (change the seed
or `n` and it lands somewhere new). That gap is the entire subject of statistics
— and none of it is meaningful until `mu` is defined.

That is the whole recap. The hard part is not "what is an estimand" in the
abstract — it is *which* estimand a real design is actually chasing, and whether
it is the one you wanted.

# What an estimand actually is {#sec-what}

## From research question to estimand {#sec-rq}

Every estimand starts life as a **research question** — usually a single
sentence in your introduction: *"Does flooding make people more supportive of
climate policy?"* That sentence is not yet an estimand. It does not say *whose*
support, measured against *what* counterfactual, *averaged how*. The estimand is
what you get when you force the research question to become precise enough that a
method could, in principle, answer it.

This research-question→estimand step is the most consequential — and most
skipped — move in a design. A loosely worded research question is compatible with
*several* estimands (effect on the treated? on everyone? on the people a policy
would actually reach?), and they need not agree. Choosing among them is a
substantive commitment about what you want to know, not a technicality to settle
later. Almost everything else in this chapter is really about doing that step
honestly: keeping the estimand faithful to the research question, instead of
letting the question quietly drift toward whatever the method happens to deliver.

## The Lundberg–Johnson–Stewart anatomy

Lundberg, Johnson, and Stewart (2021, *ASR*) give the most useful working
decomposition. A **theoretical estimand** has exactly three parts, and naming
them is most of the battle:

1. **A unit-level quantity.** What is the causal (or descriptive) quantity for a
   *single* unit? Usually a contrast of potential outcomes, $Y_i(1) - Y_i(0)$ —
   "this person's outcome if treated minus if untreated." This is never observed
   for any unit (the fundamental problem of causal inference); the switching
   equation $Y_i = D_i Y_i(1) + (1-D_i)Y_i(0)$ only ever shows you one side.
2. **A target population.** *Over whom* are we summarising? Everyone? The
   treated? Compliers? Residents of flooded communities? This is the part that
   silently changes most often.
3. **An aggregation rule.** *How* do we summarise the unit-level quantities over
   that population — a simple mean, a weighted mean, a quantile, a difference of
   subgroup means? This is the least intuitive of the three and the most often
   left implicit; @sec-aggregation takes it apart.

Write those three down and you have stated a theoretical estimand. Skip any one
and you have hand-waved. Notice that "ATE" only fixes parts 1 and 3 (a
difference in potential outcomes, simple-averaged); the *population* is left to
context — which is exactly why "the ATE" is ambiguous more often than people
admit.

::: {.callout-tip appearance="simple"}
**The three questions to ask of any result.** *What is the unit-level quantity?
Whose average is this? What population is doing the averaging?* If you cannot
answer all three about a paper's headline number, the estimand is not yet
specified — no matter how precise the standard error looks.
:::

## Three worked anatomies (and the questions they answer)

The fastest way to make the anatomy habitual is to watch the *population* slot
change while everything else stays put. Each example below starts from a
one-sentence research question and ends with a named estimand.

- **Research question:** *"Does a get-out-the-vote call make people vote?"*
  *Unit quantity:* registered voter $i$'s turnout if called minus if not called.
  *Target population:* all registered voters in the city the campaign covers.
  *Aggregation:* simple mean. → **ATE** of the call for that city's electorate.
- **Research question:** *"Did the 2014 wage-floor increase change employment at
  the firms it bound?"* *Unit quantity:* firm $j$'s headcount if the floor rose
  minus if it had not. *Target population:* firms actually bound by the increase
  (those paying below the new floor). *Aggregation:* mean over bound firms. →
  **ATT** — note the population is *the treated*, by construction.
- **Research question:** *"Would extending free school meals to currently-ineligible
  families improve attendance?"* *Unit quantity:* child $i$'s attendance if newly
  enrolled minus if not. *Target population:* children *not currently* eligible.
  *Aggregation:* simple mean. → **ATU** — the population is the people the policy
  has *not yet* reached, which is exactly why this is the hardest one to identify.

Same three-part skeleton each time; the research question dictates *whose* effect,
and that single choice is the difference between an ATE, an ATT, and an ATU. Write
the question down first and the population slot usually fills itself.

## The aggregation rule: "averaged how?" {#sec-aggregation}

All three anatomies above quietly used the *same* aggregation — a simple,
unweighted mean. That is the slot people skip, because the mean feels automatic
rather than chosen. It isn't. The same unit-level contrast over the same
population can be summarised in several ways, and they answer different questions:

- **Unweighted mean** — the default behind ATE/ATT/…: every unit counts equally,
  $E[\delta_i]$. The right choice when you genuinely want the population-average
  effect and no unit deserves more weight than another.
- **Weighted mean** — units count differently, $\sum_i w_i\,\delta_i$. The weights
  encode *what you want the number to represent*: design/sampling weights to map a
  non-representative sample onto a target population (@sec-sampling); overlap or
  propensity weights to emphasise the region where treated and control are
  comparable (@sec-restrict); policy weights to count people in proportion to whom
  a decision would actually touch. Beware that OLS supplies its *own* implicit
  weights (@sec-varweight) — rarely the ones you would have picked.
- **Quantile treatment effects (QTE)** — the effect on the *quantiles of the
  outcome distribution* rather than its mean: how the 10th or 90th percentile of
  $Y$ shifts. A programme can leave the mean untouched while lifting the bottom of
  the distribution, or raise the average only by helping those already doing well.
  (A subtlety: the QTE compares the *marginal* quantiles under treatment and
  control, $Q_\tau(Y(1)) - Q_\tau(Y(0))$ — it is *not* "the effect for the person
  at the $\tau$-th percentile," which would need the unobserved joint distribution.)
- **Difference vs. ratio (the scale)** — a mean *difference* $E[Y(1)]-E[Y(0)]$ and
  a *ratio* $E[Y(1)]/E[Y(0)]$ (risk ratio, odds ratio) are different aggregations
  of the same outcomes. "No effect" looks the same on both, but "constant effect"
  does not: effects that are constant on the difference scale are heterogeneous on
  the ratio scale, and vice versa. The scale is part of the estimand, not a
  reporting afterthought.
- **Other functionals** — a *difference of subgroup means* (do liberals move more
  than conservatives?) is itself an aggregation choice, and it is the estimand
  behind the **interaction (moderation) analysis** so common in political science:
  a treatment × moderator term does not estimate a single average effect but
  exactly this *difference in CATEs* across groups. Worth saying out loud, because
  it is easy to run the interaction and report the coefficient without noticing
  that the target has shifted from "the effect" to "the *gap between* two effects."
  The tempting "share who benefit," $P(\delta_i > 0)$, is another such functional —
  but (unlike the mean) it generally cannot be recovered without assumptions about
  who-gains-what.

::: {.callout-tip appearance="simple"}
## Why this slot is less intuitive than "for whom"
Changing the *population* changes the **set** you average over — easy to picture.
Changing the *aggregation* changes the **summary functional** applied to the same
set, which is more abstract. The intuition that fixes it: two programmes with
*identical* average effects can do completely different things — one nudging
everyone slightly, the other helping half and harming half. Same mean, opposite
policies. If your question is about *who ends up where*, not just the centre of
mass, the simple mean is the wrong aggregation no matter how cleanly you have
named the population.
:::

## Theoretical vs. empirical (statistical) estimand

Keep two layers distinct:

- The **theoretical estimand** is the substantive target — defined in the world,
  in terms of potential outcomes over a population. It exists before any data or
  identification strategy.
- The **empirical (statistical) estimand** is the function of *observable* data
  distributions that you claim equals the theoretical estimand *if your
  identifying assumptions hold*. "Difference in conditional means given the
  back-door set," "the Wald ratio," "the DiD contrast" — these are empirical
  estimands.

Identification is the bridge between the two. The empirical estimand can be
crystal clear (a well-defined functional of the data) while the theoretical
estimand it is supposed to represent is vague or contested. That gap —
*"the statistical object is clean, but what it is an estimand* of *is unclear"* —
is one of the most common things worth saying in a referee report.

## Descriptive estimands have an anatomy too {#sec-descriptive}

Not every estimand is a potential-outcome contrast. A *descriptive* estimand —
the gender gap in turnout, the share of a legislature's bills that are amended,
the trend in partisan affect — still has the full LJS anatomy: a **unit-level
quantity** that is simply a measured attribute rather than a $Y_i(1)-Y_i(0)$
difference; a **target population** to average over; and an **aggregation rule**
(mean, share, quantile, between-group difference). What changes is *identification*.
There is no counterfactual to defend, so the work is entirely about **measurement
and sampling** — does the variable capture the concept, and does the sample
represent the population? — not about confounding or assignment.

The discipline is identical: "the gender gap in turnout" is under-specified until
you name *whose* turnout (which population), *measured how*, and *summarised how*.
A precise descriptive estimand is no less demanding than a causal one; it just
fails in different ways — measurement error and sampling bias rather than
confounding. Much of @sec-sampling applies to descriptive and causal estimands
alike, because both depend on the same target-population slot.

## Same design, different estimand — because the population moved {#sec-popmoved}

Hold the data-generating process and the estimator completely fixed. Change only
*whose* effect you are averaging, and you have a different estimand:

::: {.contrast}
**Contrast 1 — same question shape, different target population.**
A job-training program is evaluated with an experiment among *volunteers*.

- *Estimand A (ATE on volunteers):* average effect for the population that
  signed up. Answers "did the program help the people who took it?"
- *Estimand B (ATE on the eligible population, including non-volunteers):*
  answers "what if we made it universal?"

Same trial, same regression. But B requires the effect for people who chose not
to enrol — exactly the people the trial never treated. If those who benefit most
self-select in (selection on gains), A > B and reporting A as if it were B
overstates the case for expansion. The estimator did not change; the *target
population* did, and with it the estimand.
:::

```{=html}
<div class="est-quiz" id="q-anatomy">
<p class="est-stem"><strong>Checkpoint.</strong> A paper reports "the effect of
attending a job fair on employment, estimated by comparing attendees to
non-attendees with rich controls." The authors call it "the ATE." On the LJS
anatomy, which part is most clearly under-specified?</p>
<label class="est-opt"><input type="radio" name="q-anatomy" data-correct="false" data-why="The unit-level contrast (employed if attend vs. if not) is actually fine here — it's a standard binary potential-outcome difference."> The unit-level quantity</label>
<label class="est-opt"><input type="radio" name="q-anatomy" data-correct="true" data-why="&quot;The ATE&quot; fixes the contrast and the simple-mean aggregation, but not the population. Attendees and non-attendees differ; with controls you most plausibly identify an effect on the treated (or on a covariate-overlap region), not the whole population. The target population is the unstated part."> The target population</label>
<label class="est-opt"><input type="radio" name="q-anatomy" data-correct="false" data-why="The aggregation (a simple average) is implied by &quot;ATE&quot; and is not the main ambiguity — the ambiguity is over whom."> The aggregation rule</label>
<label class="est-opt"><input type="radio" name="q-anatomy" data-correct="false" data-why="Something is under-specified: &quot;the ATE&quot; names a contrast and an averaging rule but leaves the population to context, and here treated/control differ systematically."> Nothing — it is fully specified</label>
<button class="est-btn" onclick="estCheck('q-anatomy')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

# Field guide: the estimand taxonomy {#sec-fieldguide}

Reference section. Each entry gives a plain-language definition, the minimal
notation, when it is the *right* target, the nearest confusion, and a one-line
example. The skimmable table at the end (@sec-taxotable) is the part to bookmark.

Throughout, $Y_i(1), Y_i(0)$ are unit $i$'s potential outcomes under treatment
and control, $D_i$ is treatment received, and $\delta_i = Y_i(1) - Y_i(0)$ is
the (unobservable) individual effect.

## ATE — Average Treatment Effect

**Plain language.** The average effect if you treated *everyone* versus *no one*,
across the entire population. **Notation.** $\text{ATE} = E[\delta_i] = E[Y_i(1) - Y_i(0)]$.
**Right target when.** Universal or near-universal interventions; "what if
everyone got this?"; mandates, across-the-board policy changes.
**Nearest confusion.** Reported when the design only credibly delivers ATT or
LATE. The ATE needs a valid counterfactual for *both* the treated and the
untreated — more demanding than people assume.
**One line.** *Effect of a universal basic income on labour supply for the whole
adult population.*

## ATT — Average Treatment Effect on the Treated

**Plain language.** The average effect among those who *actually got* treatment.
**Notation.** $\text{ATT} = E[\delta_i \mid D_i = 1]$.
**Right target when.** Evaluating an existing, targeted program for its current
recipients; retrospective "is this working?" questions.
**Nearest confusion.** Dressed up as ATE. Most quasi-experimental designs (DiD,
matching on the treated, synthetic control) deliver ATT *because they only build
a counterfactual for the treated*. That is a feature — but it is not the ATE.
**One line.** *Effect of a place-based subsidy on the firms that received it.*

## ATU — Average Treatment Effect on the Untreated

**Plain language.** The average effect for those who did *not* get treatment —
the effect they *would* experience if newly treated. **Notation.**
$\text{ATU} = E[\delta_i \mid D_i = 0]$.
**Right target when.** Expansion decisions: "if we extended the program to the
currently-untreated, what would happen?"
**Nearest confusion.** Assumed equal to ATT. Hardest parameter[^param] to identify
credibly, because almost every design builds counterfactuals relative to the
treated.

[^param]: *Parameter* is the statistician's word for a fixed, unknown feature of
the population — used throughout this chapter interchangeably with *estimand*: the
target you want to learn (the ATE, ATT, …), as distinct from the *estimator* that
guesses it and the *estimate* it produces (@sec-recap). "These parameters" later
on just means "these candidate estimands." Under selection on gains, ATU < ATT, so expansion disappoints relative
to the program's track record.
**How it's estimated.** Rarely as a direct target. Usually by modelling the
*treated* outcome surface and predicting it for the untreated (regression
imputation / g-computation), as the residual once ATE and ATT are known
($\text{ATE} = \pi\,\text{ATT} + (1-\pi)\,\text{ATU}$, with $\pi$ the treated
share), or by reverse matching/IPW. Each route extrapolates the treated outcome
model onto people who *chose not* to be treated — precisely where selection on
gains bites — which is why the ATU is the flimsiest member of the family.
**One line.** *What training would do for workers who currently decline it.*

## LATE — Local Average Treatment Effect

**Plain language.** The average effect among **compliers** — units whose
treatment status flips *because of* the instrument or assignment nudge.
**Notation.** $\text{LATE} = E[\delta_i \mid \text{complier}]$ (Imbens & Angrist
1994; Angrist, Imbens & Rubin 1996). In the experimental-noncompliance literature
the identical object is the **CACE** (complier average causal effect): same
estimand, different name — LATE emphasises the instrument, CACE the noncompliance.
**Right target when.** The contemplated policy operates at *the same margin* as
the instrument — i.e., it would move the same people.
**Nearest confusion.** Read as ATE. Three traps: (i) compliers are
**unidentified** — you cannot list them; (ii) **different instruments define
different complier groups**, hence different LATEs for the *same* treatment;
(iii) it says nothing about always-takers or never-takers.
**One line.** *Effect of military service on later earnings, for men who served
only because the draft lottery compelled them (Angrist).*

## CATE — Conditional Average Treatment Effect

**Plain language.** The ATE *within a subgroup* defined by covariates.
**Notation.** $\text{CATE}(x) = E[\delta_i \mid X_i = x]$.
**Right target when.** Heterogeneity and targeting questions; "who benefits
most?"; assessing whether an effect transports to *your* subpopulation.
**Nearest confusion.** Treated as free. Identification must hold *within* the
cell, cells are smaller (less power), and slicing many subgroups invites
false positives. A CATE averaged over the population's $X$ distribution is the
ATE — over a *different* distribution it is a different number (the seed of
transportability).
**One line.** *Effect of a GOTV mailer among first-time registrants specifically.*

## ITT — Intent-to-Treat

**Plain language.** The effect of *being assigned/offered* treatment, regardless
of take-up. **Notation.** $\text{ITT} = E[Y_i \mid Z_i = 1] - E[Y_i \mid Z_i = 0]$
for assignment $Z$.
**Right target when.** The policy lever *is* the offer (you can mandate
availability, not uptake); when you want the real-world effect of a program that
people can decline.
**Nearest confusion.** Conflated with the effect of treatment *received*. Under
imperfect compliance $\text{ITT} = \text{LATE} \times P(\text{complier})$ —
the ITT is *diluted* by non-take-up. Dividing by the first stage recovers LATE
(under exclusion + monotonicity).
**One line.** *Effect of mailing a vaccine voucher on infection rates, whether
or not recipients redeem it.* (Note: "ITT" is also used loosely for
*assignment-level* exposure even without a compliance story — see @sec-assign.)

## MTE — Marginal Treatment Effect

**Plain language.** The effect for the unit *just indifferent* between treatment
and control at a given level of unobserved resistance $u$. **Notation.**
$\text{MTE}(u) = E[\delta_i \mid U_i = u]$ (Heckman & Vytlacil 2005).
**Right target when.** You want to build *other* estimands: ATE, ATT, ATU, and
LATE are all weighted integrals of the MTE curve — they differ only in the
weights. Also for marginal-expansion policy ("treat a few more people").
**Nearest confusion.** Treated as exotic. It is actually the unifying object:
once you see the MTE curve, "which estimand?" becomes "which weights over $u$?"
**One line.** *How the return to college falls as you move from eager to
reluctant attendees.*

## Reference table {#sec-taxotable}

| Estimand | Population averaged over | One-line gloss | Most common misuse |
|---|---|---|---|
| **ATE** | Everyone | Effect of treating all vs. none | Claimed when only ATT/LATE is identified |
| **ATT** | The treated | Effect on actual recipients | Sold as ATE / as evidence for expansion |
| **ATU** | The untreated | Effect if you newly treated them | Assumed equal to ATT |
| **LATE** (= CACE) | Compliers (instrument-specific) | Effect for those the nudge moved | Read as ATE; ignoring instrument-dependence |
| **CATE** | A covariate subgroup | Effect within $X=x$ | Underpowered slicing; transport without warrant |
| **ITT** | The *assigned* | Effect of the offer | Confused with effect of receipt |
| **MTE** | The marginal unit at $u$ | Effect for the indifferent | Treated as untestably abstract |
| **QTE** | Everyone, distributionally | Effect at the $\tau$-th outcome quantile | Read as a mean effect; it is a different *aggregation rule* |

: Estimand quick-reference. **CACE** is the noncompliance-literature name for LATE. **QTE** changes the *aggregation* slot, not the population — a reminder that the third LJS part also defines the estimand. ATE, ATT, ATU, and LATE coincide only under strong conditions (@sec-coincide); the sample-vs-population distinction (SATE/PATE) is in @sec-sampling, and which design naturally targets which of these is in @sec-designmap. {.striped .hover}

```{=html}
<div class="est-quiz" id="q-taxo">
<p class="est-stem"><strong>Checkpoint.</strong> A judge-leniency IV study finds
that pre-trial detention reduces later employment. For which population is this
effect identified?</p>
<label class="est-opt"><input type="radio" name="q-taxo" data-correct="false" data-why="The IV says nothing about defendants who would be detained (or released) regardless of which judge they draw — those are always-takers/never-takers, outside the complier set."> All defendants (ATE)</label>
<label class="est-opt"><input type="radio" name="q-taxo" data-correct="false" data-why="ATT would be the effect on all who were detained, including those any judge would detain. IV only speaks to the margin the instrument moves."> Everyone who was detained (ATT)</label>
<label class="est-opt"><input type="radio" name="q-taxo" data-correct="true" data-why="Judge leniency is an instrument; it identifies a LATE for compliers — marginal defendants whose detention depends on which judge they happened to draw. A stricter or more lenient instrument would define a different complier set and a different LATE."> Marginal defendants whose detention depends on the judge (LATE)</label>
<label class="est-opt"><input type="radio" name="q-taxo" data-correct="false" data-why="MTE is the effect at a single point of indifference; the IV delivers an average over the complier interval, i.e., a LATE, not a single MTE point."> The single indifferent defendant (MTE)</label>
<button class="est-btn" onclick="estCheck('q-taxo')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

## From the estimand you have to the one you want {#sec-extrapolate}

A design hands you one parameter. Claiming a *different* one is always an
extrapolation, licensed only by an assumption you should state out loud. The
identity $\text{ATE} = \pi\,\text{ATT} + (1-\pi)\,\text{ATU}$ (with $\pi$ the
treated share) sits behind most of these moves.

| Have → Want | Assumption that licenses it | "What you'd have to believe" |
|---|---|---|
| **ATT → ATE** | ATU recoverable: homogeneity, or selection-on-observables for the *untreated* counterfactual so you can reweight | "The untreated would respond like comparable treated units — no selection on gains beyond what my covariates capture." |
| **ATT → ATU** | Same as above (ATU is the residual once ATE and ATT are pinned); at the limit, $\text{ATU} = \text{ATT}$ | "The people who declined treatment would gain what the takers gained." (Flatly false under Roy sorting.) |
| **LATE → ATE** | Homogeneous effects, *or* MTE extrapolation across the propensity range, *or* multiple instruments tracing the MTE curve, *or* shape restrictions giving bounds | "Compliers are not special — always-takers and never-takers have the same effect," or "I can extrapolate the MTE curve off the complier margin." |
| **LATE → ATT / ATU** | Effect equality across compliers and always-takers (for ATT) or never-takers (for ATU) | "The marginal units the instrument moved have the same effect as the inframarginal ones it never touched." |
| **ITT → LATE** | Exclusion + monotonicity (+ relevance); then $\text{LATE} = \text{ITT} / P(\text{complier})$ | "Assignment affects outcomes *only* through take-up, and nobody defies — so I may divide by the first stage." |

: Moving between estimands always costs an assumption about units the design never identified. {.striped}

The pattern is one idea: every upward move buys a broader population with an
assumption about the units the design *never* saw — the untreated, the
non-compliers, the inframarginal. Name which units you are extrapolating onto, and
the assumption that they resemble the ones you observed. This is also a
ready-made critique move (@sec-critique): when a paper's conclusion needs a wider
estimand than its design delivers, the question is simply *which row of this table
are they implicitly standing on, and is the belief in the last column credible?*

# Designs and the estimands they point at {#sec-designmap}

Each design has a *natural* estimand — the parameter it targets when its
assumptions hold. This section is about that logic **only**. It says nothing
about whether the assumptions are credible (parallel trends, exclusion,
continuity are someone else's section); the point is to read the target off the
design *before* arguing about validity.

One cut organises the rest. Designs differ in *where their identifying power comes
from*, and that source largely dictates how much freedom you have over the
estimand:

- **Experimental** (the RCT): the researcher randomises treatment, so a valid
  counterfactual exists by construction and you can target the ATE directly.
- **Observational / selection-on-observables** (matching, regression, IPW): there
  is no special source of exogenous variation; identification rests on the
  assumption that you have measured and conditioned on every confounder. Because
  conditioning *(if that assumption holds)* reconstructs the joint distribution of
  treated and untreated alike, **the estimand is a choice** — ATT, ATE, or ATU,
  set by how you weight. What you cannot choose is whether unconfoundedness holds.
  You would pick the ATT over the ATE when the question is about the people who
  actually received treatment (evaluating an existing programme, not a hypothetical
  universal rollout) — and because it is often the more *credible* choice, since
  the ATT needs only controls comparable to the treated, whereas the ATE demands
  overlap across the *entire* covariate space and so extrapolates the treated
  experience onto units unlike any treated unit.
- **Quasi-experimental** (DiD, IV, RDD): you do not assume confounding away; you
  exploit a slice of *as-if-random* variation in the world — a policy
  discontinuity, an instrument, a sharp timing. But that variation is *local*, so
  **the estimand is handed to you** and is usually local too: the complier LATE,
  the effect at the cutoff, the ATT among the treated-and-post. The very feature
  that buys credibility — a narrow window of exogenous variation — also narrows
  *whose* effect you identify.

Quasi-experimental designs are "observational" in the broad sense (no
researcher-controlled randomisation), and the line between the two families is
genuinely fuzzy; what separates them for our purposes is whether identification
leans on *conditioning* or on a *source of as-if-random variation*. The estimand
lesson: in observational designs you pick the target and then defend an assumption
about confounding; in quasi-experimental designs the design picks the target for
you, and your job is to notice *which* local population that is (the
identification-first route of @sec-derive).

| Family | Design | Natural estimand | Population it averages over |
|---|---|---|---|
| Experimental | Simple RCT | ATE (for the sampled population) | Everyone in the sample/frame |
| Experimental | RCT, noncompliance | **ITT** (offer) *and* **LATE/CACE** (receipt) | Assigned (ITT) vs. compliers (LATE) |
| Observational | Selection-on-observables (matching / regression / IPW) | ATT, ATE, or ATU — *whichever the weights target* | The treated, everyone, or the untreated, per estimator |
| Quasi-experimental | Canonical 2×2 DiD | ATT | The treated group |
| Quasi-experimental | IV | LATE | Compliers (instrument-specific) |
| Quasi-experimental | RDD (sharp) | Effect at the cutoff (a boundary CATE) | Units at the threshold |
| Quasi-experimental | RDD (fuzzy) | LATE at the cutoff | Compliers at the threshold |

: Design → natural estimand, grouped by where identification comes from. Estimand logic only — the credibility of each design's assumptions is a separate question. {.striped}

- **Simple RCT.** Randomisation gives a valid counterfactual for both arms, so
  the natural target is the ATE — but *for the sampled population* (@sec-sampling),
  not automatically a national one.
- **RCT with noncompliance.** Two legitimate targets, not one diluted one. ITT is
  the effect of the *offer*; LATE/CACE the effect of *receipt* among compliers;
  $\text{ITT} = \text{LATE} \times P(\text{complier})$.
- **Selection-on-observables.** The design itself does not fix the estimand —
  *the estimator's weights do*. Matching on the treated or Hájek IPW → ATT; full
  Horvitz–Thompson reweighting → ATE; reverse-matching the untreated → ATU. This
  is the one row where "which estimand?" is a choice you make, not a fact about
  the design.
- **2×2 DiD.** Builds a counterfactual *for the treated* from control trends → the
  ATT, and only the ATT.
- **IV.** Identifies the complier effect; a different instrument is a different
  complier set and a different LATE (@sec-fieldguide).
- **RDD.** A discontinuity in assignment identifies the effect *at the cutoff* — a
  local effect for threshold units, by construction not a population average.[^rdlabel]
  Fuzzy RD adds an IV step, localising further to compliers at the cutoff.

[^rdlabel]: Sources differ on the label: some call the sharp-RD target "the LATE
at the threshold," others "the ATE at the cutoff." Both mean the boundary effect
for units at the cutoff. I use *boundary CATE* to avoid colliding with this
chapter's LATE entry, which reserves "LATE" for the complier effect.

## TWFE and staggered adoption: a panel-data problem {#sec-twfe}

This is a *longitudinal-data* issue, separate from the cross-sectional
OLS-weighting point in @sec-varweight. With a single treatment time, two-way
fixed effects (TWFE) recovers the ATT cleanly. Under **staggered adoption +
heterogeneous effects**, it does not.

Goodman-Bacon (2021) shows TWFE is a weighted average of all the underlying 2×2
DiDs — including "forbidden" comparisons that use **already-treated** units as
controls for later-treated ones. Under heterogeneity the weights on some of these
comparisons can be **negative**, so the TWFE coefficient need not equal the ATT
and can even be opposite in sign to every unit-level effect. The fix is an
*estimand* fix, not a standard-error fix: heterogeneity-robust estimators —
Callaway–Sant'Anna, imputation (Borusyak et al.), Sun–Abraham — recover
interpretable **group-time ATTs** that you then aggregate with transparent,
non-negative weights.

The framing to keep: TWFE's problem is *which weighted average of group-time ATTs
the coefficient represents*, not whether the design is valid. State your target
as "the average of group-time ATTs, equally weighted across cohorts" (or whatever
aggregation you choose) and report the estimator that delivers it.

```{=html}
<div class="est-quiz" id="q-twfe">
<p class="est-stem"><strong>Checkpoint.</strong> A reviewer writes: "Your TWFE
event-study with staggered adoption may not estimate the ATT you claim." What is
the most precise restatement of the concern?</p>
<label class="est-opt"><input type="radio" name="q-twfe" data-correct="false" data-why="Clustering fixes standard errors, not the estimand. The concern is about which weighted average of effects the point estimate represents, not its variance."> The standard errors are wrong; cluster them</label>
<label class="est-opt"><input type="radio" name="q-twfe" data-correct="true" data-why="Exactly. Under staggered timing and heterogeneous effects, TWFE is a weighted average of 2×2 DiDs including forbidden already-treated-as-control comparisons, with possibly negative weights — so it can diverge from (even flip the sign of) the ATT. The fix is a heterogeneity-robust estimator (Callaway–Sant'Anna, imputation), not a different SE."> Under heterogeneity + staggered timing, TWFE is a weighted average with possibly negative weights, so it need not equal the ATT</label>
<label class="est-opt"><input type="radio" name="q-twfe" data-correct="false" data-why="Parallel trends is necessary but is a separate issue. Even if parallel trends holds perfectly, TWFE's weighting problem under heterogeneity remains."> Parallel trends fails, so the design is invalid</label>
<label class="est-opt"><input type="radio" name="q-twfe" data-correct="false" data-why="More controls don't address the variance-weighting / negative-weights problem, which is structural to TWFE under staggered timing."> You need more covariates</label>
<button class="est-btn" onclick="estCheck('q-twfe')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

# Where intuition usually breaks {#sec-breaks}

This is the section that matters. Definitions are easy; the failures below are
where careful researchers and sloppy ones part ways. The recurring move to watch
for: **the estimand quietly changes while the prose stays the same.** The failures
sort into three clusters: **(A)** the parameters genuinely diverge, so "effect for
whom?" has different answers; **(B)** the target population moves between design
and conclusion without anyone noticing; **(C)** the estimator quietly imposes a
population of its own.

## A — Effect for whom: why the parameters diverge {#sec-clusterA}

*The parameters diverge because treatment is not allocated at random with respect
to its own effect — so ATE, ATT, ATU and LATE answer different "for whom"
questions, and equality among them is a claim requiring evidence, not a default.*

### Coincidence only under strong conditions {#sec-coincide}

They are all equal **iff treatment effects are homogeneous** (constant
$\delta_i$) — or, more weakly, iff effect heterogeneity is uncorrelated with
treatment assignment and with compliance. In social science neither holds by
default. As the reference materials put it bluntly: *"everything affects everyone
differently."* So treat equality of these parameters as a claim requiring
evidence, never a convenience.

The selection-bias decomposition makes the wedge concrete. The naïve
difference in observed means is
$$
\underbrace{E[Y\mid D=1] - E[Y\mid D=0]}_{\text{what OLS-of-Y-on-D gives}}
= \text{ATE} + \underbrace{\big(E[Y(0)\mid D=1]-E[Y(0)\mid D=0]\big)}_{\text{selection bias}}
+ \underbrace{(1-\pi)(\text{ATT}-\text{ATU})}_{\text{heterogeneity term}} .
$$
The third term is non-zero *precisely when* ATT $\neq$ ATU — i.e., whenever the
people who got treated have systematically different effects from those who did
not. That is the norm, not the exception.

::: {.callout-warning appearance="simple"}
## Intuition trap: "homogeneity is the conservative assumption"
It is not conservative — it is a strong, usually unexamined modelling choice.
Assuming $\delta_i$ constant makes all the estimands collapse into one, which
*hides* the very question ("effect for whom?") that policy turns on. Conservative
would be to *report which population* your number is about.
:::

### Selection on gains and on costs (and the perfect doctor)

**Selection on gains (the Roy model).** People who expect to benefit most are
most likely to take treatment. Then $\text{ATT} > \text{ATE} > \text{ATU}$. The
program's track record (ATT) flatters what expansion (ATU) would deliver.
**Selection on costs.** Take-up is driven by who faces the lowest cost/barrier,
not who benefits most. If low-cost happens to mean low-benefit (e.g., proximity
instruments for schooling pull in students near the margin), the identified
effect can sit *below* the ATE. The lesson is not a fixed sign — it is that *the
direction of the wedge is a substantive question about who selects in and why.*

::: {.contrast}
**Contrast 2 — same estimand, different estimator.** You want the **ATT** of a
wage subsidy on recipient firms. You could get it by (a) difference-in-differences
against never-subsidised firms, or (b) matching subsidised firms to similar
unsubsidised ones. *Same target population, same contrast.* They differ only in
the assumption that licenses them — parallel trends (a) vs. unconfoundedness on
the matched covariates (b). If they disagree, at least one assumption is wrong;
they are not "two estimands." Contrast this with the next box, where the
*number moves because the target moved*.
:::

**Policy targeting — the perfect doctor.** When a planner allocates treatment to
those expected to benefit most, $\text{ATT} \gg \text{ATE}$, and ATU can be ~0 or
negative. Cunningham's *perfect doctor* assigns surgery vs. chemo optimally per
patient: the treatment genuinely helps (ATT $\approx$ +4.4 years) yet a naïve
comparison shows it *harming* people (SDO $\approx$ −0.4 years), because the
treated were the sickest. Optimal targeting maximises the gap between treated and
control counterfactuals — the cleaner the targeting, the more misleading the raw
comparison, and the less the program's record tells you about expanding it.

**Seeing the wedge in a simulation.** A binary treatment with **genuinely
heterogeneous effects** and **selection on gains**. ATE, ATT, ATU, and the naïve
difference in means all land on different numbers — none is "wrong," they are
*different estimands*.

```{r}
#| label: demo-ate-att
#| message: false
set.seed(2026)
n  <- 50000
u  <- rnorm(n)                       # latent "gain type"
y0 <- 0.5 * u + rnorm(n)
tau_i <- 0.2 + 0.8 * pnorm(u)        # individual effect: bigger for high-u units
y1 <- y0 + tau_i

# Selection on gains: high-gain units more likely to be treated.
p  <- plogis(1.2 * u)
d  <- rbinom(n, 1, p)
y  <- ifelse(d == 1, y1, y0)

ATE <- mean(tau_i)
ATT <- mean(tau_i[d == 1])
ATU <- mean(tau_i[d == 0])
naive <- mean(y[d == 1]) - mean(y[d == 0])

round(c(ATE = ATE, ATT = ATT, ATU = ATU,
        `naive diff` = naive), 3)
```

The naïve difference overshoots even the ATT, because high-gain units also had
higher $Y(0)$ (selection bias). ATT > ATE > ATU is the Roy-model signature. Flip
the effect to homogeneous (`tau_i <- rep(0.6, n)`) *and* switch selection off
(`p <- rep(0.5, n)`), and the four numbers collapse onto each other — that
knife-edge is the only place the taxonomy stops mattering. The wedge here is
driven by *selection on gains*, not by OLS weighting; that distinct channel is
@sec-varweight.

## B — The population moves without you noticing {#sec-clusterB}

*The target population can shift between design and conclusion through three quiet
channels — defining treatment at the assignment vs. exposure level, restricting or
reweighting the sample, and generalising to a population you did not study — none
of which is a bias to be corrected, and all of which require a substantive argument.*

### Assignment level vs. uptake/exposure level {#sec-assign}

A pervasive, under-discussed shift: *at what level is treatment defined?*

- **Assignment / offer:** ITT-type estimand. The effect of *being assigned* to
  (or living in) a treated condition.
- **Uptake / experienced exposure:** the effect of actually *receiving* the dose.

These differ whenever take-up or exposure is partial. The gap is usually framed
as compliance, but it shows up far more broadly — in community-level treatments
where some residents are barely touched, in encouragement designs, in
geographic "treatments" where exposure varies within the treated unit. Scaling
ITT up to a per-treated effect (dividing by a first stage) is only the *right*
estimand if the policy you have in mind achieves full uptake.

::: {.contrast}
**Contrast 3 — assignment vs. exposure (a community treatment).** A flood hits
some 1-km grid cells. Define treatment two ways:

- *Community assignment (ITT-style):* you live in a flooded cell. Averaged over
  all residents of flooded cells — including those whose own home stayed dry.
  This is a real, policy-relevant total community effect.
- *Direct individual exposure:* your own home was inundated. A within-community
  dose-response estimand, defined only inside flooded cells.

Same event, two estimands, different populations and different contrasts. The
community ITT is *not* a diluted version of "the real effect" — it is a
different, legitimate target (the effect of being in an affected place). Calling
the community estimate "the effect of being flooded" overstates it; calling the
individual estimate "the effect of the flood on the community" understates the
collective channel. (This is the CAD study's primary vs. secondary estimand;
see @sec-cad.)
:::

### Any restriction or weight relocalises the estimand {#sec-restrict}

Overlap-trimming is the textbook case, but the general point is broader and
matters more in political science, where formal common-support trimming is rare
and usually unremarked. *Any* move that changes which units enter the average —
trimming for overlap, dropping off-support cells, applying sampling or propensity
weights, restricting to a subsample "for robustness" — silently moves the target
population, hence the estimand. "ATT on the supported treated" is a different
claim from "ATT on the treated," and a weighted estimate targets a different
population from its unweighted twin (@sec-sampling). The discipline is one
sentence: whenever you restrict or reweight, say which sub-population the number
now refers to.

### Transportability vs. internal estimation

Two different failures that get blurred:

- **Internal estimation problem (bias):** your estimate of the chosen estimand,
  *for your sample's population*, is biased — confounding, bad controls,
  violated parallel trends.
- **Transportability problem (target population):** your estimate of the chosen
  estimand is *fine for the population you studied*, but you are using it to
  speak about a *different* population where effects may differ.

These call for different responses. Bias is fixed (if at all) by better
identification. Transportability is a substantive argument about *why effects
should or should not carry over* — there is no purely statistical fix. Mislabel
a transportability gap as bias and you will "correct" the wrong thing.

::: {.contrast}
**Contrast 4 — identified estimand vs. desired estimand.** A draft-lottery IV
cleanly identifies the **LATE** of military service for lottery compliers. Your
policy question is the **ATE** for all potential servicemembers (including
volunteers and those who would never serve). The estimate is *not biased* — it
is an excellent estimate of the LATE. It is simply not an estimate of the thing
you want. The fix is not a better instrument for the *same* design; it is either
a different design, an extrapolation argument (MTE, multiple instruments,
bounds), or honesty that the question has shifted.
:::

## C — The estimator imposes its own estimand {#sec-clusterC}

*Even with identification settled, the estimator's weighting can target a
population nobody chose — so "which estimand?" is partly a question about your
method, not only your design.*

### OLS with controls and the variance-weighted estimand {#sec-varweight}

A regression of $Y$ on a binary $D$ plus controls does not, in general, return
"the ATE" under effect heterogeneity. But the size of the problem is bounded, and
it is worth saying precisely *where it bites* — because OLS-with-controls is the
workhorse of the discipline and a blanket warning is not useful.

**Where it is (nearly) fine.** When the propensity to be treated is *constant*
across units — a clean RCT with a single assignment probability, or adjustment
that is *saturated* (treatment fully interacted with discrete covariate strata) —
regression adjustment recovers the ATE to first order. In a randomised experiment
covariates are a precision device, not an identification one; the estimand is
still the ATE. (Two caveats from the experimental-adjustment literature: additive
covariate adjustment can induce a minor finite-sample bias under heterogeneity;
Lin's fully-interacted estimator removes it and cannot hurt asymptotic precision.
For typical political-science $N$ this is second-order.)

**Where it bites.** Under *selection-on-observables with a non-constant propensity
score* — the usual observational case — the OLS coefficient on $D$ is a
**variance-weighted** average of strata-specific effects: strata where treatment
is most variable ($\hat p(X)\approx 0.5$) dominate, while near-always-treated and
near-never-treated strata contribute almost nothing. This is *not bias* — the
estimate is consistent for a weighted average — but the weights are an artefact of
the treatment-probability distribution, not a population you chose. Interacting
treatment with covariates restores a target-able estimand; full saturation removes
the weighting problem at the cost of power.

**The Słoczyński (2022) result, precisely.** Even with a single binary treatment
and no functional-form subtlety, OLS does not return the ATE under heterogeneity.
It converges to a convex combination
$$ \hat\beta_{\text{OLS}} \;\to\; (1-\rho)\,\text{ATT} + \rho\,\text{ATU}, \qquad \rho = \text{treated share}, $$
weighting ATT by the *untreated* share and ATU by the *treated* share — the
**opposite** of the ATE weights. So OLS puts *more* weight on the group that is
the *smaller* share of the sample. With 10% treated, $\hat\beta$ loads roughly 90%
on the ATT and barely sees the ATU; with a small control group it sits near the
ATU. The ATE ($\rho\,\text{ATT} + (1-\rho)\,\text{ATU}$) is recovered only at
$\rho = 0.5$ or under homogeneity.

**How big is the divergence?** Bounded by two things and no more: how far the
treated share is from one-half, and the gap $\text{ATT}-\text{ATU}$ (the size of
selection-driven heterogeneity). If either is small, so is the divergence — equal
shares, or near-homogeneous effects, and OLS $\approx$ ATE. Resist a universal
number; the wedge is study-specific. Claim "OLS departs from the ATE in proportion
to treated-share imbalance times effect heterogeneity," not "OLS is typically far
from the ATE."

::: {.callout-warning appearance="simple"}
## Verdict you can act on
Worry when **both** hold: the treated (or control) group is a small share of the
sample, *and* you have a substantive reason to expect $\text{ATT}\neq\text{ATU}$
(selection on gains, targeting). Then "the OLS coefficient" is neither ATE, ATT,
nor ATU, and you should either state which weighted average you report or switch
to a weighting/saturated estimator that targets a named parameter. Do *not* worry
much in a balanced experiment, or where effects are plausibly homogeneous in the
dimension separating treated from control. The staggered-**panel** version of this
problem (TWFE, negative weights) is separate and more severe — see @sec-twfe.
:::

# Sampling pins down the population {#sec-sampling}

The LJS anatomy has three slots; sampling decides the **second one** — the target
population — and it does so whether or not you write it down. Two estimands with
the same unit-level contrast and the same aggregation rule are *different
estimands* if they average over different populations (@sec-popmoved). Sampling is
how that population gets fixed, and it is the slot a survey experiment most often
leaves implicit.

## SATE and PATE

Hold the experiment fixed and ask: whose average?

- **SATE** — the **sample** average treatment effect,
  $\text{SATE} = \tfrac1n\sum_{i\in\text{sample}}\delta_i$: the average effect over
  *the units you actually have*.
- **PATE** — the **population** average treatment effect, $\text{PATE} = E[\delta_i]$
  over the target population the sample was meant to represent.

They coincide when the sample is a representative draw — formally, when sampling
is independent of the treatment effect $\delta_i$. They diverge whenever the units
you recruited have systematically different effects from those you did not: a
*sampling* analogue of selection on gains. What closes the gap is exactly one of
two things: (i) **probability sampling** from the target population (then the SATE
is an unbiased estimate of the PATE by design); or (ii) a **reweighting argument**
under an *ignorable-selection* assumption — selection into the sample is
independent of $\delta_i$ *conditional on observed covariates* $W$, so reweighting
to the population's $W$-distribution recovers the PATE. Route (ii) is the sampling
twin of selection-on-observables, and just as untestable.

## Probability samples vs. opt-in panels

This is where most survey-experimental papers quietly overclaim. A clean RCT
*embedded in a convenience or quota panel* (a YouGov-style quota sample, a student
pool, MTurk) has impeccable **internal** validity — random assignment identifies
an ATE *for the people in the panel*. What it does **not** identify, without
further argument, is a national PATE. The respondent pool is self-selected,
quota-filled, or platform-specific; its members can differ from the national
population on precisely the dimensions that moderate the effect.

So the **defensible estimand of a quota/opt-in survey experiment is "the ATE for
the respondent pool,"** not "the ATE for [country] adults" — unless the authors
make an explicit reweighting + ignorable-selection argument. Most papers report
something like *"data are from YouGov, 3,000 quota-sampled respondents,"* present
weighted estimates, and then discuss them as national effects without ever stating
the assumption that licenses the jump. The estimand is left implicit — which is
the recurring theme of this chapter (@sec-extrapolate): there is always a target,
and silence does not make it the national one.

- **When implicit is fine.** If the effect is plausibly *homogeneous* over the
  dimensions on which the panel differs from the population — a framing effect with
  no theoretical reason to vary by the panel's skews — the SATE–PATE gap is small
  and the elision is defensible.
- **When it is not.** If the effect is heterogeneous and the panel is skewed on a
  moderator (MTurk skews young, educated, low-trust; student samples skew on
  political sophistication), reweighting on demographics does not fix selection on
  the *moderator* unless that moderator is in $W$. Then the national claim is a
  transportability overreach (@sec-critique), not a weighting technicality.

## Weights *are* an estimand choice

Sampling weights are not a nuisance correction; **applying them changes the target
population, hence the estimand.** Unweighted, you estimate the SATE for the
achieved sample; weighted to population margins, you target the PATE for the
population those margins describe. Switching weights silently switches the estimand
— the same move as the restriction/reweighting point in @sec-restrict, in a
different costume. Report which population the weights define, and note when an
"unweighted robustness check" is therefore a *different estimand*, not merely a
less efficient one.

## A secondary, observational example

The point is not specific to surveys. In an administrative-data study — say, the
effect of a job-centre programme using full municipal registers — the *sampled
frame is the population*: you observe (near) everyone in the frame, so SATE = PATE
*for that frame*. The estimand question then collapses onto the frame's definition:
"all unemployment-insured residents of these three municipalities, 2015–2019" is
the population, and any claim beyond it (other municipalities, other years) is
transportability, not sampling. Defining the frame *is* defining the target
population — the second LJS slot — and a frame restriction (drop non-citizens, drop
short spells) relocates the estimand exactly as a sample restriction does
(@sec-restrict).

```{=html}
<div class="est-quiz" id="q-sampling">
<p class="est-stem"><strong>Checkpoint.</strong> A survey experiment fielded on a
3,000-person opt-in quota panel finds a framing effect and concludes "framing
shifts [country]'s public opinion." The authors apply demographic weights. The
cleanest statement of the estimand issue is:</p>
<label class="est-opt"><input type="radio" name="q-sampling" data-correct="false" data-why="Random assignment within the panel gives clean internal validity; the issue is not bias in estimating the panel's effect."> The randomisation is broken, so the estimate is biased</label>
<label class="est-opt"><input type="radio" name="q-sampling" data-correct="true" data-why="Right. The design identifies the SATE for the panel; the national PATE claim needs probability sampling or an ignorable-selection-given-W argument. Demographic weights only close the gap if selection into the panel is ignorable conditional on those demographics — and effect-moderators not in W (e.g. political attentiveness) break that."> It identifies the effect for the panel (a SATE); the national claim needs probability sampling or ignorable selection given the weighting variables</label>
<label class="est-opt"><input type="radio" name="q-sampling" data-correct="false" data-why="Demographic weights target population *margins* on the weighting variables; they do not guarantee the PATE unless selection is ignorable given exactly those variables, which is an assumption, not an automatic fix."> Applying demographic weights makes it the national PATE automatically</label>
<label class="est-opt"><input type="radio" name="q-sampling" data-correct="false" data-why="A bigger panel shrinks the standard error around the SATE; it does nothing about the sample-to-population gap, which is a population question, not a precision one."> A larger panel would fix the problem</label>
<button class="est-btn" onclick="estCheck('q-sampling')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

# Estimands in the design pipeline {#sec-pipeline}

## MIDA: the estimand is the Inquiry

DeclareDesign (Blair, Coppock, and Humphreys) frames any design as four
components, **M-I-D-A**:

- **M**odel — your beliefs about how the world works (units, potential outcomes,
  unobservables).
- **I**nquiry — *the estimand*: the question asked of the model.
- **D**ata strategy — sampling and assignment: how data are produced.
- **A**nswer strategy — the estimator applied to the data.

The estimand is the **I**. Everything else is machinery for answering it. The
discipline this enforces: you can write down the Inquiry *before* you have data,
and you can check — in simulation — whether your Answer strategy actually
recovers it under your Model. A design "works" when the Answer strategy returns
the Inquiry; "diagnosands" like bias and power are *properties of the M-I-D-A
quadruple*, not of the estimator alone.

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "The pipeline. The estimand (Inquiry) is chosen up front; the estimator only recovers it when the identification assumptions linking the empirical to the theoretical estimand hold."
flowchart LR
  RQ["Research question<br/><i>(substantive)</i>"] --> EST["<b>Estimand</b><br/>theoretical: unit quantity<br/>+ population + aggregation"]
  EST --> EMP["Empirical estimand<br/><i>function of observables</i>"]
  EMP -. "identifying<br/>assumptions" .-> ID{"hold?"}
  ID -- "yes" --> ER["Estimator<br/><i>(answer strategy)</i>"]
  ID -- "no" --> BIAS["estimate of<br/>something else"]
  ER --> E["Estimate<br/><i>(a number + CI)</i>"]
  classDef target fill:#d4efdf,stroke:#1e8449,color:#145a32;
  classDef warn fill:#fadbd8,stroke:#c0392b,color:#922b21;
  class EST target;
  class BIAS warn;
```

## How much of an estimand is "precisely stating what you're doing" vs. "the thing you want to know"?

Both, and the tension is the point. A lot of estimand work is *bookkeeping*:
forcing yourself to write the population and aggregation rule so a reader knows
exactly what your number refers to. But it is not *only* bookkeeping, because
the act of writing it down repeatedly reveals that the thing you can identify is
**not** the thing you wanted — that your design hands you an ATT when your
conclusion needs an ATU, or a LATE for an odd complier group when you wanted a
population effect. The estimand is where the **research question** and the
feasible design are forced to confront each other. Stating it precisely is how
you *notice* the mismatch — the moment you see that the quantity you can identify
is not the one your research question asked for — instead of papering over it.

Make the division of labour explicit:

- the **estimand is chosen** — it is a decision about what you want to know;
- the **estimator is a procedure** — a rule applied to data;
- the **estimate is the realised output** — one number from that rule;
- a method delivers a given estimand **only if its assumptions hold**. "DiD
  estimates the ATT" means "DiD estimates the ATT *if parallel trends holds*." If
  it doesn't, you have neither the ATT nor any other tidy parameter — just a
  biased number.

## Declaring an inquiry in code

You can write the estimand as executable code — which forces all three LJS parts
into the open. The snippet below states the **Inquiry** (the ATE) directly. It
is shown both as idiomatic DeclareDesign and as a base-R twin, so the document
renders even where DeclareDesign is unavailable.

```{r}
#| label: demo-declare-inquiry
#| message: false
#| warning: false
if (requireNamespace("DeclareDesign", quietly = TRUE)) {
  library(DeclareDesign)

  design <-
    declare_model(
      N = 1000,
      U = rnorm(N),
      potential_outcomes(Y ~ 0.3 * Z + 0.8 * U * Z + U)
    ) +
    # The ESTIMAND, stated explicitly as the average unit-level contrast:
    declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
    declare_assignment(Z = complete_ra(N, prob = 0.5)) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(Y ~ Z, inquiry = "ATE")

  set.seed(2026)
  draw_estimands(design)          # the inquiry's true value for one draw
} else {
  # base-R twin: same population, same inquiry computed by hand
  set.seed(2026)
  N <- 1000; U <- rnorm(N)
  Y0 <- U; Y1 <- 0.3 + 0.8 * U + U
  data.frame(inquiry = "ATE", estimand = mean(Y1 - Y0))
}
```

The value worth internalising: `declare_inquiry()` is the line where you commit
to *what you want to know*, separately from `declare_estimator()`, where you
commit to *how you will guess it*. Keeping those two lines distinct in your head
— even when you are not using DeclareDesign — is most of the skill.

```{=html}
<div class="est-quiz" id="q-mida">
<p class="est-stem"><strong>Checkpoint.</strong> In MIDA terms, a colleague says
"my estimator is unbiased, so my design is fine." What is missing?</p>
<label class="est-opt"><input type="radio" name="q-mida" data-correct="true" data-why="Unbiasedness is a property of the Answer strategy *relative to a specified Inquiry under a Model*. &quot;Unbiased&quot; is meaningless until you say unbiased *for what estimand*, and a design can be unbiased for an Inquiry nobody wanted."> "Unbiased for *which* Inquiry?" — bias is defined only relative to a stated estimand</label>
<label class="est-opt"><input type="radio" name="q-mida" data-correct="false" data-why="More data improves precision but does not tell you whether the estimator targets the right estimand; an estimator can be unbiased for the wrong target at any N."> Nothing — more data will settle it</label>
<label class="est-opt"><input type="radio" name="q-mida" data-correct="false" data-why="A bigger model isn't the issue; the issue is that &quot;unbiased&quot; presupposes a fixed Inquiry that hasn't been named."> A more flexible functional form</label>
<label class="est-opt"><input type="radio" name="q-mida" data-correct="false" data-why="Clustering addresses variance, not the prior question of which estimand unbiasedness is being claimed against."> Clustered standard errors</label>
<button class="est-btn" onclick="estCheck('q-mida')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

# How to derive an estimand in practice {#sec-derive}

Three routes. Most good empirical work uses the second; most *published* work
quietly uses the first; the third is where you end up when feasibility bites.
Know all three and know their failure modes.

## Route 1 — Canonical-first

**Move.** Reach for an off-the-shelf parameter — ATE, ATT, LATE — and fit your
study to it.
**Strength.** Fast, communicable, comparable across literatures. Often genuinely
correct: if you ran a clean experiment on your population of interest, "the ATE
for this population" *is* your estimand and there is no need to overthink it.
**Failure mode.** Lazy category-fitting: you call it "the ATE" because that is
the box your method came in, not because the whole population is the target.
The label then does identification work it has not earned.
**At the workshop table.** *"You've labelled this the ATE — but your design only
ever builds a counterfactual for the treated. Why is the ATT not the honest
name here?"*

## Route 2 — Question-first / estimand-first

**Move.** Start from the substantive research question or policy question. Name
the unit-level quantity, the target population, and the aggregation rule *before*
choosing a method. Then ask which design could identify *that*.
**Strength.** Keeps the answer tethered to the question. You discover early
whether the thing you care about is even identifiable, instead of after the
fieldwork.
**Failure mode.** Defining a beautiful estimand no feasible design can recover,
then either abandoning the project or quietly swapping in whatever you *can*
estimate while keeping the original prose.
**At the workshop table.** *"My estimand is the ATU — the effect on
currently-untreated municipalities, because the policy question is expansion. I
know my DiD only gives the ATT, so let me be explicit about the extrapolation I
need."*

## Route 3 — Identification-first / backwards

**Move.** Start from what the design credibly identifies — "I have a clean
instrument," "I have a discontinuity" — and read off the estimand it implies.
**Strength.** Honest about feasibility; it is how you avoid claiming things your
data cannot support. Sometimes the identified estimand turns out to be exactly
what matters.
**Failure mode.** Letting the tail wag the dog: the LATE for a weird complier
group becomes "the effect of X" because that is what the instrument gave you,
and the introduction is quietly rewritten to pretend that was the question all
along.
**At the workshop table.** *"What this RD identifies is the effect at the
threshold, for units near the cutoff. Before I sell it as the program's effect, is
the cutoff population the one the decision is actually about?"*

::: {.callout-tip appearance="simple"}
## A checklist for deriving an estimand from a study idea
1. **Unit quantity.** Write the single-unit contrast in words, then symbols.
   ("Person $i$'s turnout if canvassed minus if not.")
2. **Population.** Name *exactly whom* you average over. Resist "the population"
   — which one? Everyone, the treated, compliers, a covariate region?
3. **Aggregation.** Simple mean? Weighted by what? A quantile? A subgroup
   difference?
4. **Assignment vs. exposure.** Is the estimand defined at the level of *offer/
   assignment* or *received dose*? Are they the same here?
5. **Feasible counterpart.** What empirical estimand (functional of observables)
   would equal this *if* assumptions held? Name the assumptions.
6. **Gap check.** Is the feasible estimand the same population/contrast as step
   2–4? If not, say so now — that gap is your paper's central honesty.
7. **Policy match (if relevant).** Does the estimand's population coincide with
   the population the decision is about? If not, you have a transportability
   argument to make, not just an estimate to report.
:::

# Writing estimands in papers {#sec-writing}

This section is meant to be lifted directly into a draft.

## The estimand sentence: a recipe

A good estimand sentence names, in one or two sentences, the four things a
reader needs to reconstruct your target:

> **[unit-level contrast]** for **[target population]**, **[aggregation rule]**,
> where treatment is defined at the level of **[assignment / exposure]**.

You will not always need the fourth clause, but include it whenever assignment
and uptake/exposure can differ.

## Templates

- **Clean ATE (experiment on the population of interest):**
  *"Our estimand is the average effect of [treatment] on [outcome] among
  [population], defined as the population mean of the individual treatment
  effect $Y_i(1)-Y_i(0)$."*
- **ATT (observational/quasi-experimental):**
  *"We estimate the average treatment effect on the treated: the mean effect of
  [treatment] on [outcome] for the units that received it, $E[Y_i(1)-Y_i(0)\mid
  D_i=1]$. We do not identify the effect for never-treated units."*
- **LATE (IV):**
  *"Our estimand is the local average treatment effect — the effect of
  [treatment] for compliers, i.e. units induced into treatment by [instrument].
  Compliers are not the full population, and a different instrument would define
  a different complier group."*
- **ITT / assignment-level:**
  *"Our estimand is the intent-to-treat effect of [offer/assignment]: the effect
  of being [assigned/offered], regardless of take-up. This is the policy-relevant
  quantity because the lever available to policymakers is the offer, not uptake."*
- **Overlap-localised ATT:**
  *"After trimming to the region of common covariate support, our estimand is the
  ATT for the supported sub-population of treated units ($X\%$ of treated units
  fall off-support and are excluded; we describe them in Table N)."*

## Implicit estimands are still estimands

The templates above make something explicit that papers usually leave unsaid:
*"we estimate the effect of X on Y"* already commits to an estimand, because every
estimator presupposes a target. You cannot run a regression that is the answer to
*no* question. So "stating the estimand" is rarely inventing one from nothing — it
is **surfacing the one the method and assumptions already imply** (DiD → ATT, IV →
LATE, OLS-with-controls → a variance-weighted average; see @sec-designmap).

Leaving it implicit is fine when the design admits only one sensible target — a
clean RCT on the population of interest, where "the effect" can only mean the ATE
for that population. It becomes a problem exactly when several targets are
compatible with the same sentence, which is the common case in observational and
survey work. The test: *could a careful reader write down a different LJS anatomy
that is equally consistent with your prose?* If yes, the estimand needs saying out
loud; if no, the implicit version is honest shorthand.

## How many estimands should a paper declare?

Count carefully, because the axes multiply. A design usually pins **one contrast
structure** (one population × contrast × aggregation), but each *outcome* is a
separate estimand and each *moderation* analysis adds more — so a typical paper
carries one primary, several secondary-outcome estimands, and a few
CATE/interaction estimands. Robustness checks are **not** new estimands — same
target, different estimator — *unless* they trim or reweight and quietly move the
population (@sec-restrict), which is a target-shift hiding in a robustness table.

The discipline is ranking, not a magic number:

- **Declare exactly one primary estimand** — one outcome, one contrast, one
  population, one aggregation — in a single sentence, and power for it.
- **Label the rest as secondary**, and keep confirmatory separate from
  exploratory; more estimands means more multiplicity risk, not more findings.
- **Treat moderation as a distinct, underpowered class.** Pre-specify one or two
  moderators rather than sweeping; interactions need far larger samples than main
  effects (Gelman's rule of thumb, ≈16× when the interaction is half the size).
  Report it as the *gap-between-effects* estimand (@sec-aggregation), flagged
  exploratory unless pre-registered.

## Good / acceptable / sloppy

::: {.phrasing}
| Verdict | Phrasing | Why |
|---|---|---|
| **Good** | "We estimate the ATT of receiving the subsidy on firm survival, for firms that received it between 2010–2015; we do not estimate the effect for non-recipients." | Names contrast, population, scope; flags what is *not* identified. |
| **Acceptable** | "We estimate the effect of the subsidy on firm survival using difference-in-differences." | Method implies ATT to an expert reader, but leaves the population implicit. Passable; not best practice. |
| **Sloppy** | "We estimate the effect of the subsidy on firm survival." | Effect for whom? Contrast against what? Assignment or receipt? The reader cannot reconstruct the target. |
| **Sloppy & misleading** | "We find the subsidy raises survival, suggesting expanding it would help more firms." | Slides from an ATT to an implicit ATU claim — exactly the population the design cannot speak to. |
:::

## Worked examples from real traditions

**Coppock, *Persuasion in Parallel* — clarity as a model.** Coppock's estimand
is unusually well-specified: the *per-message* treatment effect on attitudes,
estimated across many parallel experiments, with explicit attention to whether
effects are homogeneous *across subgroups* (do liberals and conservatives move
in parallel?). The contribution is partly substantive and partly *estimand
hygiene*: by stating the contrast and the populations (subgroups) precisely, the
"parallel" claim becomes a testable statement about CATEs being equal across
groups, not a vibe. Worth emulating: he says what the effect is *for whom* and
then asks whether it is the *same* for whom-else.

**Angrist & Imbens-style IV-LATE — honesty about the complier population.** In
Angrist's draft-lottery work and the Imbens–Angrist (1994) framework, the
estimand is explicitly the LATE: the return to military service *for men whose
service was induced by the lottery*. The papers are careful that this is **not**
the effect for volunteers or for those who would never serve. The template to
steal: state the instrument, state that the estimand is the complier effect,
and state that the complier population is instrument-specific and not
enumerable.

**An LJS-style anatomy — write the three parts down.** For a study of the effect
of a cash transfer on school enrolment: *unit quantity* = child $i$'s enrolment
if the household receives the transfer minus if not; *target population* =
children in eligible poor households in the study districts; *aggregation* =
simple mean (optionally re-weighted to the national eligible population, which
would be a *different, transported* estimand). Spelling these out makes the
national-vs-local choice a visible decision rather than an accident of wording.

::: {.contrast}
**Contrast 5 — same estimate label, different target population.** Two RD studies
of a means-tested scholarship both report "the effect of the scholarship on
graduation." Study A's cutoff sits at the 20th income percentile; Study B's at
the 60th. Both are valid RDs. Both identify an effect *at their own threshold* —
two different CATEs for two different sub-populations. Reported under the same
label "the effect," they look like a replication (or a contradiction); read as
estimands, they are simply *different local effects* and may both be right. "Same
label, different population" is one of the most common sources of fake
disagreement in a literature.
:::

# How to critique other researchers' estimands {#sec-critique}

A reviewer's and seminar-goer's toolkit. The goal is to diagnose *which kind* of
problem you are looking at, because the kinds call for different responses and
different language.

## Questions to ask of any empirical paper

1. **Can I state the estimand in LJS terms from what's written?** If not, the
   first problem is *specification*, before anything else.
2. **Whose effect is this?** Treated, untreated, compliers, a covariate region,
   the assigned?
3. **Does the method's natural estimand match the stated one?** (DiD→ATT,
   IV→LATE, OLS/TWFE→variance-weighted average.)
4. **Does the population stay fixed from design to conclusion?** Watch the
   introduction and the policy paragraph especially.
5. **Assignment or exposure?** Is the headline the effect of the *offer* or of
   *receipt*, and is the prose consistent?
6. **Did overlap-trimming or sample restrictions move the population?**
7. **Is a within-study estimate being transported to a different population?**

## The five diagnoses

Name the failure precisely — it changes what you ask the author to do.

| Diagnosis | Tell-tale sign | What you ask for |
|---|---|---|
| **Unclear estimand** | You cannot reconstruct population/contrast from the text | "Please state the estimand: effect for whom, contrast against what." |
| **Wrong estimand** | The stated target doesn't answer the stated question | "The question is about expansion (ATU), but you identify the ATT." |
| **Shifted target population** | Population changes between design and conclusion | "The design is on compliers; the conclusion is about everyone." |
| **Estimator/estimand mismatch** | Method's natural estimand ≠ claimed estimand | "TWFE under staggered timing isn't the ATT you claim." |
| **Transportability overreach** | Clean internal estimate, wrong population for the claim | "This is well-identified *for site A*; the policy claim is about site B." |

## Red flags and category errors

- **"The effect of X"** with no population named — the default sloppiness.
- **ATT evidence, ATE/ATU rhetoric** — "the program works, so let's scale it";
  make them name the extrapolation assumption it rests on (@sec-extrapolate).
- **"We estimate the ATE using OLS with controls"** under obvious heterogeneity —
  almost certainly a variance-weighted average, not the ATE (@sec-varweight).
- **A single LATE doing population-level work**, with no acknowledgement that the
  complier group is instrument-specific.
- **Standard-error talk used to answer estimand questions** — precision is not
  relevance; an exquisitely tight CI around the wrong estimand is still the wrong
  estimand.
- **"Robustness" trimming** that quietly localises the estimand without saying so.

::: {.callout-tip appearance="simple"}
## Workshop / referee language you can reuse
- *"I think the design identifies X, but the paper is written as if it identifies Y."*
- *"This reads as an ATT claim with ATE rhetoric — can you restrict the policy implication to the treated?"*
- *"Is the problem here bias in estimating your chosen estimand, or that the chosen estimand is not the one your conclusion is about?"* (Internal vs. transportability — force the distinction.)
- *"Which population is doing the averaging in your main number?"*
- *"Your empirical estimand is clear; I'm unsure what theoretical estimand it corresponds to."*
- *"A different instrument would give a different complier group — how special is this LATE?"*
:::

```{=html}
<div class="est-quiz" id="q-diagnose">
<p class="est-stem"><strong>Checkpoint.</strong> A paper runs a clean RCT in one
NGO's clinics, estimates the ATE on that patient population, and concludes "this
intervention should be adopted nationally." The most precise critique is:</p>
<label class="est-opt"><input type="radio" name="q-diagnose" data-correct="false" data-why="If it's a clean RCT, the within-study estimate of the ATE is likely unbiased. The problem is not internal bias."> The estimate is biased by confounding</label>
<label class="est-opt"><input type="radio" name="q-diagnose" data-correct="false" data-why="Estimator/estimand mismatch would mean the method doesn't target the claimed parameter. An RCT does target the ATE for its sample. That's not the issue."> The estimator targets the wrong parameter (mismatch)</label>
<label class="est-opt"><input type="radio" name="q-diagnose" data-correct="true" data-why="Right. The estimand is well-identified *for the NGO's patient population*; the national claim asserts it transports to a different population where effects may differ. This is a transportability/target-population problem, requiring a substantive argument, not a statistical correction."> Transportability overreach: well-identified for the study population, extrapolated to a different one</label>
<label class="est-opt"><input type="radio" name="q-diagnose" data-correct="false" data-why="The estimand *is* stated (ATE for the clinic population). The problem isn't that it's unstated — it's that the conclusion silently swaps the population."> The estimand is unclear / unstated</label>
<button class="est-btn" onclick="estCheck('q-diagnose')">Check answer</button>
<div class="est-feedback"></div>
</div>
```

# Running example: the CAD coastal-flooding study {#sec-cad}

This section applies the chapter to a specific study — a post-event,
cross-sectional, quasi-experimental design measuring the effect of community
flood exposure on political attitudes in coastal Denmark. External readers can
skip it.

## Research question and primary estimand

**Research question.** *Does living in a community that floods change residents'
political attitudes — for example, their support for climate-adaptation policy,
trust in government, or risk perceptions?* As @sec-rq insists, that sentence is
not yet an estimand: it fixes neither *whose* attitudes nor the contrast against
which "change" is measured. Forcing it to be precise yields the target below.

The **community-level ITT effect**: the average effect of *living in a 1×1 km
grid cell that flooded* on individual political attitudes, in the population of
**adult permanent residents of coastal flood-risk cells**. In LJS terms:

- **Unit quantity:** resident $i$'s attitude if their cell flooded minus if it
  did not.
- **Target population:** adult permanent residents of coastal flood-risk
  communities (cells that contain at least one resident inside an official SDFI
  flood-risk polygon).
- **Aggregation:** a population mean (with design weights for descriptive
  quantities; the causal contrast is the cell-level treatment coefficient).

Run through the estimand-sentence recipe of @sec-writing, that becomes one
declarative line:

> *Our estimand is the average effect of one's 1×1 km community of permanent
> residence flooding versus not flooding, on political attitudes, for adult
> permanent residents of coastal flood-risk cells, expressed as a population mean
> — where treatment is defined at the level of community **assignment** (living in
> a flooded cell), not individual flood exposure.*

"ITT" here is meant in the **assignment** sense (@sec-assign) — assignment to a
flooded *community* — **not** the compliance ITT of an experiment with
non-compliers. The docs are explicit about this distinction, which is exactly
the kind of clarification this chapter argues for.

## What is the treatment assignment level?

The **cell** (1×1 km kvadratnet). Treatment is `Flooded_j` at the cell level;
SEs cluster at the cell; the binding constraint on power is $J_T$, the number of
flooded *cells*, not the number of respondents per cell. This is a community
treatment, so the assignment-vs-exposure distinction is live and deliberate: a
resident of a flooded cell whose own home stayed dry is still *treated* under
the primary estimand.

## What is the estimand under heterogeneity? (a candid wrinkle)

The estimation doc is admirably honest: with effect heterogeneity, OLS with
segment fixed effects returns a **variance-weighted ITT** (@sec-varweight), not
necessarily the ATT for the flooded population. They therefore *label* $\tau$ as
a variance-weighted ITT and report a design/IPW-weighted ATT-on-the-flooded as a
robustness check. This is the @sec-varweight problem named and managed rather
than hidden — a model of good practice.

## Why permanent residents but not summer houses?

The target population is **permanent residents**, a substantive choice, not just
a data limitation. The community-exposure *mechanism*
requires day-to-day embedding in the affected place: local services, neighbours,
visible disruption, recovery politics, municipal response. A summer-house owner
registered elsewhere experiences a *different exposure channel* — property
damage and financial loss — which is a different unit-level quantity over a
different population. (There is also an operational reason: CPR samples people at
their permanent address, so summer-house-only addresses are not sampleable. But
the docs lead with the *estimand* rationale, correctly.)

## "Estimate on summer houses, claim about year-round homes" — a transportability trap (hypothetical)

The study *excludes* summer-house owners from the estimand, so it does **not**
make this error — but the hypothetical is the cleanest illustration of
@sec-breaks. Suppose you *did* estimate effects among flooded summer-house owners
(property-damage channel) and then spoke about *permanent residents* of flooded
communities. That is **not bias** in estimating the original estimand — it is a
**target-population / transportability problem**: a clean effect for population A
(second-home owners, property channel) used to claim about population B
(year-round residents, community channel) where the mechanism and the effect may
differ. The fix is not a better estimator; it is either restricting the claim or
arguing substantively that the channels coincide (they plausibly do not).

Relatedly, the docs treat **control contamination** — permanent-resident
controls who own a *flooded* summer house elsewhere — as a measured robustness
issue, *not* an automatic attenuation. That is the right call: the bias sign is
only "conservative" under a same-sign assumption about second-home exposure,
which is itself a substantive claim to be checked, not assumed.

## Frames A vs. C vs. D — when the control frame changes the estimand

This is @sec-popmoved in the wild. The frames differ in *which residents count*
on each side — whether you keep the whole 1×1 km cell or crop it down to the
in-polygon (mapped-risk) slice — and that cropping choice moves the estimand:

| Frame | Treated cells | Control cells | Estimand it targets | Why this status |
|---|---|---|---|---|
| **C** (primary) | whole cell | whole cell | the **community** effect — exposed vs. unexposed community | crops neither side, so the same covariate "lens" applies to both at full sample size — matches the primary community estimand |
| **A** (sensitivity) | whole cell | in-polygon slice | *asymmetric*: full-community treated vs. mapped-risk controls; leans toward individual mapped-risk exposure | crops controls only; the lens mismatch leaves A's controls systematically lower-lying and wetter than treated, so it is kept only as an exposure-cleanliness check |
| **D** (rejected) | in-polygon slice | in-polygon slice | a clean **individual** mapped-risk estimand — a *different* target from C | crops both sides; restores a symmetric lens but switches the target to an individual-level estimand and deletes 18–60% of treated clusters — the docs instead fix the lens with C (un-crop controls), not by cropping treated |

: What each frame crops, the estimand it implies, and why it has the status it does. {.striped}

**Why is Frame B not here?** There is a Frame B, dropped for a *measurement* flaw
rather than an estimand shift: it qualified any cell whose geometry merely
*intersected* a flood polygon, sweeping in whole urban cells with no flood-exposed
residents (62.9% of B's cells contained zero in-polygon residents). That is a
contaminated operationalisation of the *same* community target, not a different
target — so it belongs to the measurement-error family, not this section's
estimand-choice family.

That is precisely the discipline of this chapter: a sampling-frame choice is not
just an efficiency knob, it can silently redefine *whose effect you are
estimating* — even as a different-looking frame (B) leaves the estimand intact and
fails on measurement instead.

## Community ITT vs. individual direct-exposure: what switching the language changes

The **secondary estimand** is the *individual dose-response*: within flooded
cells, the effect of *direct* physical/economic exposure (property damage,
flood depth) on attitudes, with cell fixed effects. Switching from community-ITT
language to
individual-direct-exposure language changes all three LJS parts at once:

- **population:** all residents of flooded cells → only directly-exposed
  residents *within* flooded cells;
- **contrast:** flooded-community vs. not → more-exposed vs. less-exposed
  *within the same cell*;
- **identification:** as-good-as-random flooding across cells → within-cell
  conditional independence given floor level, dwelling value, etc.

$\delta$ (direct exposure) is the effect *given that you live in a flooded
community* — it does **not** recover $\tau$ (the total community effect), and
conflating the two in prose would be a textbook assignment-vs-exposure slip.

## Overlap and the localised estimand

The pre-invitation common-support screen keeps
all treated cells but restricts the *control* draw to the supported region —
which, as @sec-breaks notes, **localises the estimand to the overlap region**.
For the few unsupported treated cells (inland "surprise floods" with no coastal
counterfactual), the docs pre-specify the handling and note that dropping them
makes the estimand "the ATT for the *supported* sub-population." Again: the
target-population consequence is named, not buried.

# Conclusion: be the one watching {#sec-conclusion}

Every paper invites a discussant, even when none is in the room — someone reading
the design section with one finger on the text and an eyebrow raised, asking *who,
exactly, this number is about*. That habit is the whole chapter compressed into a
reflex. You read a results table and your first question is no longer "is this
significant?" but "significant **for whom**, averaged how?" You write a design
section and you say the population out loud, because you have seen how quietly it
moves — from assigned to exposed, from compliers to everyone, from one clinic's
patients to a national mandate. You referee a draft with the language ready: *the
design identifies X, but the paper is written as if it identifies Y.*

Every failure in this chapter was the same mistake in a different coat: someone
answered "effect how?" without first settling "effect for whom?", and the prose
never noticed when the target underneath it shifted. The taxonomy, the LJS
anatomy, the assignment-versus-exposure and bias-versus-transportability
distinctions are not decoration — they are how you hold the target still long
enough to see whether your design can actually hit it. The point is not more
notation. It is clearer targets.

None of this guarantees the right answer. It guarantees you are answering a
question you could state — a lower bar than it sounds, and one most papers clear
by accident rather than by design. So name the estimand, name the population it is
about, and ask whether the conclusion is still talking about the same one. The
estimand quietly changes while the prose stays the same — unless someone is
watching. **Be the one watching.**

# Final self-test and further reading {#sec-selftest}

## Consolidated self-test

Six questions spanning the chapter. Immediate feedback; no peeking at the
explanations until you have committed.

```{=html}
<div class="est-quiz" id="s1">
<p class="est-stem"><strong>1.</strong> "DiD estimates the ATT" is best read as:</p>
<label class="est-opt"><input type="radio" name="s1" data-correct="false" data-why="It is not unconditional. DiD targets the ATT only when parallel trends holds; otherwise it estimates a biased number that is neither ATT nor any tidy parameter."> An unconditional fact about DiD</label>
<label class="est-opt"><input type="radio" name="s1" data-correct="true" data-why="Correct: a method targets an estimand *conditional on its identifying assumptions*. &quot;DiD estimates the ATT&quot; = &quot;DiD estimates the ATT if parallel trends holds.&quot;"> A claim conditional on parallel trends holding</label>
<label class="est-opt"><input type="radio" name="s1" data-correct="false" data-why="DiD's natural target is the ATT (it builds a counterfactual for the treated), not the ATE, which would require the untreated counterfactual too."> Evidence that DiD targets the ATE</label>
<button class="est-btn" onclick="estCheck('s1')">Check</button><div class="est-feedback"></div>
</div>

<div class="est-quiz" id="s2">
<p class="est-stem"><strong>2.</strong> You trim 12% of treated units for lack of
common support. The estimand is now:</p>
<label class="est-opt"><input type="radio" name="s2" data-correct="false" data-why="It is no longer the ATT for *all* treated units — you dropped some, and they may be systematically distinctive."> Still the ATT for all treated units</label>
<label class="est-opt"><input type="radio" name="s2" data-correct="true" data-why="Right. Trimming localises the estimand to the supported sub-population; report it as such and describe who was dropped."> The ATT for the supported sub-population</label>
<label class="est-opt"><input type="radio" name="s2" data-correct="false" data-why="The ATE requires the untreated counterfactual too; trimming for overlap does not turn an ATT into an ATE."> The ATE for the full population</label>
<button class="est-btn" onclick="estCheck('s2')">Check</button><div class="est-feedback"></div>
</div>

<div class="est-quiz" id="s3">
<p class="est-stem"><strong>3.</strong> Two IV studies of the same treatment use
different instruments and report different effects. The first hypothesis should be:</p>
<label class="est-opt"><input type="radio" name="s3" data-correct="false" data-why="At least one being biased is *possible*, but with different instruments the leading explanation is that they identify LATEs for different complier groups."> One of them must be biased</label>
<label class="est-opt"><input type="radio" name="s3" data-correct="true" data-why="Correct: different instruments define different complier populations, so different LATEs are expected even if both are internally valid. The divergence is informative about heterogeneity."> They estimate different LATEs for different complier groups</label>
<label class="est-opt"><input type="radio" name="s3" data-correct="false" data-why="Averaging two LATEs over different, unknown complier populations does not yield a meaningful population parameter."> Average the two to get the ATE</label>
<button class="est-btn" onclick="estCheck('s3')">Check</button><div class="est-feedback"></div>
</div>

<div class="est-quiz" id="s4">
<p class="est-stem"><strong>4.</strong> A community-level "flooded cell" treatment,
averaged over all residents of flooded cells, is most precisely:</p>
<label class="est-opt"><input type="radio" name="s4" data-correct="true" data-why="Yes — an assignment-level (ITT-style) community effect. It is a legitimate target, distinct from the individual direct-exposure effect, not a diluted version of it."> An assignment-level community effect (ITT-style)</label>
<label class="est-opt"><input type="radio" name="s4" data-correct="false" data-why="The individual direct-exposure effect is a different estimand — defined only for directly-exposed residents within flooded cells, with a within-cell contrast."> The effect of personally being flooded</label>
<label class="est-opt"><input type="radio" name="s4" data-correct="false" data-why="There is no instrument/compliance structure here; this is assignment-level exposure, not a complier LATE."> A LATE for complier residents</label>
<button class="est-btn" onclick="estCheck('s4')">Check</button><div class="est-feedback"></div>
</div>

<div class="est-quiz" id="s5">
<p class="est-stem"><strong>5.</strong> "We estimate the ATE with OLS controlling
for confounders," effects plainly heterogeneous. The honest label is usually:</p>
<label class="est-opt"><input type="radio" name="s5" data-correct="false" data-why="OLS with controls recovers the ATE only under homogeneity (or very special weighting). Under heterogeneity it generally does not."> The ATE, as stated</label>
<label class="est-opt"><input type="radio" name="s5" data-correct="true" data-why="Correct: under heterogeneity, OLS returns a variance-weighted average (Słoczyński shows it mis-weights ATT and ATU) — generally not the ATE."> A variance-weighted average that is generally not the ATE</label>
<label class="est-opt"><input type="radio" name="s5" data-correct="false" data-why="The ATU is the effect on the untreated specifically; OLS does not target it either."> The ATU</label>
<button class="est-btn" onclick="estCheck('s5')">Check</button><div class="est-feedback"></div>
</div>

<div class="est-quiz" id="s6">
<p class="est-stem"><strong>6.</strong> "Is the problem bias in estimating your
chosen estimand, or that the chosen estimand isn't the one your conclusion is
about?" This question separates:</p>
<label class="est-opt"><input type="radio" name="s6" data-correct="false" data-why="Both sides of the question can arise with any sample size; it is not about precision/power."> Large samples from small samples</label>
<label class="est-opt"><input type="radio" name="s6" data-correct="true" data-why="Exactly: it forces the distinction between an internal estimation problem (bias for the chosen estimand) and a target-population/transportability problem (right estimate, wrong population for the claim)."> Internal-estimation problems from target-population/transportability problems</label>
<label class="est-opt"><input type="radio" name="s6" data-correct="false" data-why="It is about *what* is being estimated vs. *how well*, not about which software or estimator package you used."> Frequentist from Bayesian estimation</label>
<button class="est-btn" onclick="estCheck('s6')">Check</button><div class="est-feedback"></div>
</div>
```

## Further reading

- **Lundberg, Johnson & Stewart (2021), "What Is Your Estimand?", *American
  Sociological Review*** — the unit-quantity / target-population / aggregation
  anatomy, and the theoretical-vs-empirical estimand distinction. The single
  most useful paper for this chapter.
- **Imbens & Angrist (1994), "Identification and Estimation of Local Average
  Treatment Effects", *Econometrica*** — the LATE theorem.
- **Angrist, Imbens & Rubin (1996), *JASA*** — IV through potential outcomes; the
  complier / always-taker / never-taker / defier taxonomy.
- **Coppock, *Persuasion in Parallel* (2023)** — a model of estimand clarity and
  of testing whether CATEs are equal across subgroups.
- **Blair, Coppock & Humphreys, *Research Design in the Social Sciences*** (and
  the **DeclareDesign** R package) — MIDA, the estimand as Inquiry, and
  diagnosing designs by simulation.
- **Cunningham, *Causal Inference: The Mixtape*** — selection-bias decomposition,
  the perfect-doctor example, accessible LATE treatment.
- **Huntington-Klein, *The Effect*** — variance-weighting intuition; "the way you
  estimate the effect matters too."
- **Słoczyński (2022), *Review of Economics and Statistics*** — how OLS with a
  binary treatment mis-weights ATT and ATU.
- **Goodman-Bacon (2021), *Journal of Econometrics*** — the TWFE decomposition,
  forbidden comparisons, and negative weights under staggered timing.
- **Heckman & Vytlacil (2005), *Econometrica*** — the MTE and the unification of
  treatment-effect parameters as weighted integrals of one curve.

## Next time you read or write a design section

::: {.callout-tip appearance="simple"}
**A pocket checklist.**

*Reading:* (1) What is the **research question**, and does the stated estimand
actually answer *it*? (2) Can I state the estimand in LJS terms from the text?
(3) Whose effect is it? (4) Does the method's natural estimand match the claim?
(5) Does the population stay fixed from design to conclusion? (6) Is this a bias
problem or a target-population problem?

*Writing:* (1) Write the unit contrast in words, then symbols. (2) Name the exact
population. (3) State the aggregation. (4) Specify assignment vs. exposure if
they can differ. (5) Name what you do **not** identify. (6) If you trimmed or
restricted, say which sub-population the estimand now refers to.
:::
